From physical principles to ???

### 2016-04-20 2015CompleteDerivation/completeDerivation.tex
Another pass at classical Hamiltonian section

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we now explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space $\mathcal{C}$ of the composite system and the state space $\mathcal{S}$ of its parts. In general, this is quite a complicated thing to do, that requires intimate knowledge of the system at hand. So we simplify our problem and study an idealized case: one where the components are homogeneous and infinitesimal. What we'll find is that under the additional assumption that the system is infinitesimally reducible (i.e. its state is equivalent to the state of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian particle mechanics.

@@

As we study the system under evolution, we'll also want to study state changes. For example, if a stable mixture of gas expands, we'll want to know the difference between the initial and final distributions. That is: $\Delta\mathcal{c}=\hat{\mathcal{c}}-\mathcal{c}$. Note that such state difference may not describe a physical state: it may remove some material from one location to add it somewhere else. Yet, these state changes are still physically distinguishable objects that provide a configuration for the same homogeneous material, therefore we extend the state space $\mathcal{C}$ to include state differences. All combined, this gives us the structure of an abelian group.

@@

We claim $\mathcal{C}$ is an additive group. As we want to describe changes during the evolution (i.e. $\hat{\mathcal{c}} = \mathcal{c} +\Delta \mathcal{c}$) we introduce an inverse $- : \mathcal{C} \rightarrow \mathcal{C}$ such that $\mathcal{c} + ( - \mathcal{c}) = 0 \forall \mathcal{c} \in \mathcal{C}$. Such inverse may introduce objects that do not properly represent a physical state, but a state change. A change of a physically distinguishable object is itself physically distinguishable therefore $\mathcal{C}$ is still a Hausdorff and second countable topological manifold as per \ref{statedef}. We'll still call $\mathcal{C}$ state space committing an abuse of terminology.
\end{justification}

@@

As we combine the state of finite parts, we sum the distributions over particle states $\rho(\mathcal{s})=\rho_1(\mathcal{s})+\rho_2(\mathcal{s})$: the amount of material in the composite state is the sum of the amount of material of the parts. We can also increase or decrease the amount of material by a constant factor $\rho_1(\mathcal{s})=a\rho_2(\mathcal{s})$. This gives us the structure of a real vector space that is isomorphic to the space of continuous functions.

@@

We claim $\mathcal{C}$ is isomorphic to a subspace of $C(\mathcal{S})$ as a vector space over $\mathbbm{R}$. Let $\rho:\mathcal{S} \rightarrow \mathbbm{R}$ be a map returning the amount of material for each state $\mathcal{s} \in \mathcal{S}$. $\rho \in C(\mathcal{S})$ by \ref{continuous_map}. Let $G \subset C(\mathcal{S})$ be the functions that correspond to physically meaningful configurations. To each $\mathcal{c} \in \mathcal{C}$ corresponds one and only one $\rho_\mathcal{c} \in G$ as the system is infinitesimally reducible. Let $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$, then $\rho_\mathcal{c}=\rho_{\mathcal{c}_1}+\rho_{\mathcal{c}_2}$ as the amount of material of the composition of two states is the sum of the individual amounts. Let $\mathcal{c}_1=\tau(a)\mathcal{c}_2$, then $\rho_{\mathcal{c}_1}=a\rho_{\mathcal{c}_2}$ as $\tau(a)$ increases the amount of material by the factor $a$. $\mathcal{C}\cong G$ as vector spaces. $G$ is a subspace of $C(\mathcal{S})$.

Note that not all continuous functions correspond to physically meaningful states: we must make sure they describe a finite amount of material.

\subsection{Coordinate invariant densities}

For each classical composite state will also tell us the total amount of material, once we fixed unit (e.g. mole, kg, ...). That is, there will be a functional $\Lambda : \mathcal{C} \rightarrow \mathbbm{R}$ that tells us how much gas we have for a particular gas configuration. As we combine parts, the total amount of material will be the sum of their parts. As we increase or decrease by a constant factor the amount of material will increase or decrease by the same factor. That is: $\Lambda(a_1 \mathcal{c}_1 + a_2 \mathcal{c}_2) = a_1 \Lambda(\mathcal{c}_1) + a_2 \Lambda(\mathcal{c}_2)$. For a proper state, one that is not the difference of two states, we will expect a positive amount of material. So, if a density $\rho(\mathcal{S})$ is positive everywhere, then the total amount is also positive. This makes $\Lambda$ a positive linear functional.

Being able to tell the total amount is not enough. Experimentally, what we do is measure the amount of material for a region of phase space. For example, the amount of gas within a particular region of space for a particular range of momentum. In general, given an outcome of our physical process (i.e. an open set in the topology for $\mathcal{S}$), each composite state should tell us the amount of material associated with it (i.e. contained in the open set of states). For each $U \subseteq \mathcal{S}$ we'll have a positive linear functional $\Lambda_U : \mathcal{C} \rightarrow \mathbbm{R}$ whose value is always bounded.

The ability to associate finite amount of material with sets of states allows us to define finite regions of $\mathcal{S}$ and compare them. Consider a region of position and momentum in phase space. If we are able to spread a finite amount of material into a uniform distribution, then we know we have a finite region. And the density will give us an indication of how big the region is: if we double the region, the density will half. In other words: because we are describing densities over $\mathcal{S}$, we are able to give a unique way to measure the size of sets of $\mathcal{S}$, a metric $d\mu$ and our positive linear functionals become integrals: $\Lambda_U (\mathcal{c}) : \int_U \rho_{\mathcal{c}} d \mu$.

In mathematical terms, we can use the Riesz representation theorem for linear functionals to give $\mathcal{S}$ the structure of a measure space, with a suitable Borel $\sigma$-algebra and measure $\mu$.

@@

\begin{assump}[Classical hamiltonian assumption]\label{classical}
	The evolution of the classical homogeneous system under study is deterministic, reversible and reducible.
\end{assump}

>>>>

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we now explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space $\mathcal{C}$ of a composite system and the state space $\mathcal{S}$ of its parts. In general, this is quite a complicated thing to do, that requires intimate knowledge of the system at hand. So we simplify our problem and study the system is made of infinitesimal homogeneous parts. What we'll find is that under the additional assumption that the system is infinitesimally reducible (i.e. its state is equivalent to the state of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian mechanics.

@@

As we study the system under evolution, we'll also want to study state changes. For example, if a stable mixture of gas expands, we'll want to know the difference between the initial and final distributions. That is: $\delta\mathcal{c}=\hat{\mathcal{c}}-\mathcal{c}$. Note that such state difference may not describe a physical state: it may remove some material from one location to add it somewhere else. Yet, these state changes are still physically distinguishable objects that provide a configuration for the same homogeneous material, therefore we extend the state space $\mathcal{C}$ to include state differences. All combined, this gives us the structure of an abelian group.

@@

We claim $\mathcal{C}$ is an additive group. As we want to describe changes during the evolution (i.e. $\hat{\mathcal{c}} = \mathcal{c} +\delta \mathcal{c}$) we introduce an inverse $- : \mathcal{C} \rightarrow \mathcal{C}$ such that $\mathcal{c} + ( - \mathcal{c}) = 0 \forall \mathcal{c} \in \mathcal{C}$. Such inverse may introduce objects that do not properly represent a physical state, but a state change. A change of a physically distinguishable object is itself physically distinguishable therefore $\mathcal{C}$ is still a Hausdorff and second countable topological manifold as per \ref{statedef}. We'll still call $\mathcal{C}$ state space committing an abuse of terminology.
\end{justification}

@@

As we combine the state of finite parts, we sum the distributions over particle states $\rho(\mathcal{s})=\rho_1(\mathcal{s})+\rho_2(\mathcal{s})$: the amount of material in the composite state is the sum of the amount of material of the parts. We can also increase or decrease the amount of material by a constant factor $\rho_1(\mathcal{s})=a\rho_2(\mathcal{s})$. This gives us the structure of a real vector space that is isomorphic to a subspace of continuous functions.

@@

We claim $\mathcal{C}$ is isomorphic to a subspace of $C(\mathcal{S})$ as a vector space over $\mathbbm{R}$. $\forall \mathcal{c} \in \mathcal{C} \exists ! \rho_{\mathcal{c}}:\mathcal{S} \rightarrow \mathbbm{R}$ returning the amount of material for each state $\mathcal{s} \in \mathcal{S}$. $\rho \in C(\mathcal{S})$ by \ref{continuous_map}. Let $\varrho : \mathcal{C} \rightarrow C(\mathcal{S})$ such that $\varrho(\mathcal{c}) \mapsto \rho_\mathcal{c}$. 
As the system is reducible, two distinct composite state must represent different distributions. $\forall \mathcal{c_1}, \mathcal{c_2} \in \mathcal{C}, \mathcal{c_1} \neq \mathcal{c_2} \implies \varrho(\mathcal{c_1}) \neq \varrho(\mathcal{c_1})$. $\varrho$ is injective. $\varrho$ is a bijection between $\mathcal{C}$ and $\varrho(\mathcal{C})$. Let $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$, then $\varrho(\mathcal{c})=\varrho(\mathcal{c}_1)+\varrho(\mathcal{c}_2)$ as the amount of material of the composition of two states is the sum of the individual amounts. Let $\mathcal{c}_1=\tau(a)\mathcal{c}_2$, then $\varrho(\mathcal{c}_1)=a \varrho(\mathcal{c}_2)$ as $\tau(a)$ increases the amount of material by the factor $a$. $\varrho$ is a homomorphism. $\mathcal{C}\cong \varrho(\mathcal{C}) \subseteq C(\mathcal{S})$ as vector spaces.
\end{justification}

\subsection{Invariant densities}

In the case where the topology of $\mathcal{S}$ is not discrete, the distribution $\rho_\mathcal{c}$ is a density which is not something we directly measure. What we do measure experimentally are finite amounts of material. For example, the amount of fluid in a particular volume within a particular range of momentum specified in some units (e.g. mass, kg, ...). This means that given a composite state $\mathcal{c}$ and a set $U \subseteq \mathcal{S}$ of particle states compatible with an outcome of a process, we must be able to tell the amount of material that we will find associated with that outcome. That is for each open set $U \subseteq \mathcal{S}$ there will be a functional $\Lambda_U : \mathcal{C} \rightarrow \mathbbm{R}$.

As we combine parts, the total amount of material will be the sum of their parts. As we increase or decrease by a constant factor the amount of material will increase or decrease by the same factor. That is: $\Lambda_U(a_1 \mathcal{c}_1 + a_2 \mathcal{c}_2) = a_1 \Lambda_U(\mathcal{c}_1) + a_2 \Lambda_U(\mathcal{c}_2)$. For a proper state, one that is not the difference of two states, we will expect a positive amount of material. So, if a density $\rho_\mathcal{c}(\mathcal{S})$ is positive everywhere, then the total amount is also positive. This makes $\Lambda_U$ a positive linear functional. As the amount of material we will measure is always finite, the functional applied to any composite state $\mathcal{c}$ over any set $U$ will be finite.

The ability to associate finite amount of material with sets of states allows us to define finite regions of $\mathcal{S}$ and compare them. Consider a region of position and momentum in phase space. If we are able to spread a finite amount of material into a uniform distribution, then we know we have a finite region. And the density will give us an indication of how big the region is: if we double the region, the density will half. In other words: because we are describing densities over $\mathcal{S}$, we are able to give a unique measure $\mu$ for the size of sets of $\mathcal{S}$, a metric $d\mu$ and our positive linear functionals become integrals: $\Lambda_U (\mathcal{c}) : \int_U \rho_{\mathcal{c}} d \mu$.

This intuitive picture is formalized mathematically by the Riesz representation theorem for linear functionals, which gives $\mathcal{S}$ the structure of a measure space, with a suitable Borel $\sigma$-algebra and measure $\mu$.

@@

\begin{assump}[Classical Hamiltonian assumption]\label{classical}
	The system under study is composed of an infinitesimally reducible homogeneous material and each part undergoes deterministic and reversible evolution.
\end{assump}
